Acceleration of plague outbreaks in the second pandemic\\\emph{Electronic \supp}

### 2020-06-19 supp.Rnw
two clarifications in generation interval section

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next epidemic cycle becomes infectious is:

\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}

From \cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}

Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

>>>>

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.
